Light and growth form interact to shape stomatal ratio among British angiosperms

### 2017-10-21 ms/ms.tex
resubmitted

$^1$ Biodiversity Research Centre and Botany Department, University of British Columbia, Vancouver, British Columbia V6T 1Z4, Canada \\
\\
\textit{Author for correspondence:} \\
\textit{Christopher D. Muir} \\
\textit{Tel:} +17782284851 \\
\textit{Email:} chrisdmuir@gmail.com \\
University of British Columbia \\
6270 University Blvd. \\
Vancouver, BC, Canada \\
V6T 1Z4 \\
\\
Short title: Shedding light on stomatal evolution\\
\\
Word count: \\
Summary: 190\\
Introduction: 1179\\
Materials and Methods: 1176\\
Results: 433\\
Discussion: 1504\\
Acknowledgement: 24\\
3 Figures and 1 Table, 1 Supplemental Figure

\linenumbers
\setstretch{\stretchy}
%\pagenumbering{gobble}

@@

\begin{itemize}
	\item In most plants, stomata are located only on the abaxial leaf surface (hypostomy), but many plants have stomata on both surfaces (amphistomy). High light and herbaceous growth form have been hypothesized to favor amphistomy, but these hypotheses have not been rigourously tested together using phylogenetic comparative methods.
	\item I leveraged a large dataset including stomatal ratio, Ellenberg light indicator value, growth form, and phylogenetic relationships for 372 species of British angiosperms. I used phylogenetic comparative methods to test how light and/or growth form influence stomatal ratio.
	\item High light and herbaceous growth form are correlated with amphistomy, as predicted, but they also interact; the effect of light is pronounced in therophytes (annuals) and perennial herbs, but muted in phanerophytes (shrubs and trees). Interestingly, amphistomy and stomatal density evolve together in response to light, suggesting coordinated selection on this trait combination.
	\item Comparative analyses of British angiosperms reveal two major insights into physiological evolution. First, light and growth form interact to shape variation in stomatal ratio; amphistomy is advantageous in high light, but mostly for herbs. Second, strong coevolution of adaxial stomatal density and light tolerance indicates that amphistomy is an important adaptation to optimally coordinate light acquisition with gas exchange. These results advance our understanding of why stomatal ratio evolves and its potential as a functional trait for paleoecology and crop improvement.
\end{itemize}

@@

% MARKER (still revising, adding predictions here): Specifically, I predicted that species with faster life histories, especially therophytes (annuals), would have greater \els~than species with slower life histories, especially phanerophytes (shrubs and trees).

The final question is important for addressing whether amphistomy is part of a coordinated syndrome of traits that promote higher photosynthetic rate, as both the light and growth form hypotheses assume. If evolved increases in stomatal ratio are mediated by shifting abaxial stomata to the adaxial surface, holding total stomatal density constant, then the overall increase in CO$_2$ diffusion would be small. In contrast, if amphistomy evolves by increasing adaxial stomatal density while holding abaxial density constant, then \textit{total} stomatal density must increase as well. Evolutionary coordination of amphistomy and high stomatal density would reinforce one another, increasing CO$_2$ supply to chloroplasts more than changes in either trait would in isolation. Understanding selection on coordinated traits can explain the evolution of major functional trait axes and syndromes.

@@

Overall, $\mathrm{SR_{even}}$ increased with \el, but there was a significant interaction ($\Delta\mathrm{AIC} > 2$, Table~\ref{table:srAIC}) between Raunki\ae r life form and \el~(Fig.~\ref{fig:SR-lf}). When classified based on plant habit, growth form interacted with \el less ($\Delta \textrm{AIC} = 2.4$; Fig.~\ref{fig:SR-hf}). Raunki\ae r life form explained variation in stomatal ratio better than habit (lower AIC; Table~\ref{table:srAIC}), therefore we focus hereafter on those analyses. Both life form and \el~significantly increased model fit, though \el~had a markedly larger effect on model AIC (Table~\ref{table:srAIC}). The significant interaction is caused by different slopes between life forms. Among life forms with the overall greatest \el~(therophytes, hemicryptophytes, and chamaephytes, see Fig.~\ref{fig:lf-light}), there was a strong positive relationship between \el~and $\mathrm{SR_{even}}$. Parametrically bootstrapped 95\% confidence intervals for the slope did not overlap zero (Fig.~\ref{fig:SR-lf}). The slope was weakly positive or not significantly different from zero in the most shade-adapted life forms (geophytes and phanerophytes), albeit the patterns were distinct in these groups. There were both hypostomatous ($\mathrm{SR_{even}} \approx 0$) and amphistomatous ($\mathrm{SR_{even}} \approx 1$) geophytes, but these were distributed across \el s. In contrast, phanerophytes were nearly always hypostomatous regardless of \el.

@@

The weak or nonsignificant relationship between \el~and stomatal ratio in geophytes and phanerophytes suggests that in some cases amphistomy may not reliably indicate open habitat without further information. For example, perhaps amphistomatous geophytes from partially shaded habitats are spring ephemerals, so they  experience high light during their growth phase, but this has not been tested. Likewise, phanerophytes (mostly tall trees) are almost always hypostomatous \citep[see also][]{Muir_2015}. Most British phanerophytes are tall, hypostomatous trees, but the exceptions are telling. For example, the most amphistomatous phanerophyte in this dataset is \textit{Brassica oleracea}, a short-statured biennial that has more in common physiologically with hemicryptophytes than other phanerophytes. The other amphistomatous phanerophytes in this data set (\textit{Populus nigra} and \textit{Lavatera arborea}) are fast-growing pioneer species.

Finally, phylogenetic information should improve inferences about paleoclimates because there is appreciable phylogenetic signal in stomatal ratio. The phylogenetic half-life of stomatal ratio evolution, after accounting for \el~and Raunki\ae r life form, is $\mathrm{log}(2) / \alpha = 1.5$ my (see Table~\ref{table:srAIC} for maximum likelihood estimates of $\alpha$, the return rate in the Ornstein-Uhlenbeck model of trait evolution). This lag time may indicate that evolving to the `optimum' is constrained by the shape of the fitness landscape \citep{Muir_2015} or that other unmeasured factors which affect stomatal ratio have some phylogenetic signal. Regardless of the mechanism, this fact means that researchers may be able to use data from closely related species to improve paleoenvironment reconstruction. Despite there being phylogenetic signal, residual phylogenetic variation in stomatal ratio at the broad phylogenetic scale encompassed by British angiosperms should be at stationarity. The variance in stomatal ratio, after accounting for \el~and Raunki\ae r life form, was indistinguishable at stationarity under an Ornstein-Uhlenbeck process (see Results). This may not be the case for younger clades that have radiated in the past few million years.

\subsection*{Why does adaxial stomatal density control stomatal ratio?}

Variation in stomatal ratio is determined primarily by evolution of adaxial stomatal density and is coordinated with increases in total leaf stomatal density summed across both surfaces. Note here that I am referring only to evolutionary variation in stomatal ratio among species; different processes may mediate within species variation or plastic responses. Phylogenetic analyses show that changes in stomatal ratio and total stomatal density, especially in response to \el, are predominantly mediated by changes in adaxial stomatal density. This highly nonrandom pattern among British angiosperms mirrors evolutionary changes wrought by domestication \citep{Milla_etal_2013}; crop species tend to have higher adaxial stomatal density than their wild relatives.

@@

I refer to the second hypothesis as the `coordination' hypothesis. In this scenario, ecological conditions such as high light select for both increased total stomatal density and for amphistomy because these traits work well in coordination with one another. For example, if stomatal density were very high on a hypostomatous plant, then CO$_2$ would be more strongly limited by the mesophyll. Adding a second parallel pathway for diffusion by developing stomata on both surfaces would restore a more optimal balance between stomatal and mesophyll limitations. Conversely, there would be little benefit to amphistomy when total stomatal density is low because CO$_2$ diffusion is strongly limited by stomatal resistance, and therefore photosynthetic rate is not sensitive to changes in mesophyll diffusion mediated by stomatal ratio. A related prediction is that increased atmospheric CO$_2$ may select for reduced stomatal ratio and density primarily by decreasing adaxial stomatal density, but this has not been well tested \citep[but see][]{Woodward_Bazzaz_1988}.

@@

\begin{table}[ht]
 \setstretch{\stretchy}
  \topcaption{Interaction beween species' Ellenberg light indicator value (\el) and Raunki\ae r lifeform predict stomatal ratio ($\mathrm{SR_{even}}$). I compared phylogenetic linear models using the Akaike Information Criterion (AIC), where AIC = $2k - 2 \mathrm{log}(\mathcal{L})$. $k$ is the number of model parameters and $\mathcal{L}$ is the model likelihood. Given a set of candidate models, the difference in AIC between a model and the lowest AIC ($\Delta$AIC) indicates the relative fit of competing models. The correlation coefficient $r^2$ is another indicator of model fit. $\alpha$ and $\sigma^2$ are the return rate and diffusion parameters of the Ornstein-Uhlenbeck model of trait evolution. }
  \begin{center}
  \begin{tabular}{@{} l c c c c c c c @{}}
  \toprule
  Model: $\mathrm{SR_{even}} \sim$ &  $\alpha$ & $\sigma^2$ &  $r^2$ & $k$ &  log($\mathcal{L}$) &  AIC &  $\Delta$AIC \\
  \midrule  
    \el~$\times$ Raunki\ae r lifeform
  & 0.46
  & 0.068
  & 0.34
  & 12
  & -33.3
  & 90.6
  & 0 \\
    \el~$\times$ growth form
  & 0.46
  & 0.07
  & 0.32
  & 12
  & -38.2
  & 100.3
  & 9.8 \\
    \el~+ Raunki\ae r lifeform
  & 0.47
  & 0.072
  & 0.32
  & 8
  & -40.3
  & 96.5
  & 6 \\
     \el~+ growth form
  & 0.51
  & 0.08
  & 0.31
  & 8
  & -43.3
  & 102.7
  & 12.1 \\
    Raunki\ae r lifeform
  & 0.34
  & 0.067
  & 0.15
  & 7
  & -79.2
  & 172.4
  & 81.8 \\
    growth form
  & 0.35
  & 0.069
  & 0.13
  & 7
  & -82.5
  & 178.9
  & 88.4 \\
    \el
  & 0.64
  & 0.108
  & 0.26
  & 4
  & -59.3
  & 126.6
  & 36.1 \\
    null
  & 0.29
  & 0.067
  & 0
  & 3
  & -107.5
  & 221
  & 130.5 \\
	\bottomrule
	\end{tabular}
	\label{table:srAIC}
\end{center}
\end{table}

@@

\begin{figure}[ht]
\centerline{\includegraphics[width=\textwidth]{figures/figure_phylo-lifeform.pdf}}
\caption{Phylogenetic diversification of stomatal ratio follows growth form and light tolerance. At the center is the phylogenetic tree for 372 species of British angiosperms. For each species, the green wedges indicate Raunki\ae r life form and the orange wedges indicate \el. The outer circle indicates the stomatal ratio for each species. As shown in the legend above, greater stomatal ratio means stomata are more evenly distributed across both leaf surfaces; lower stomatal ration means that most stomata are on the lower surface.} 
\label{fig:phylo-lifeform}
\end{figure}

\begin{figure}[ht]
\centerline{\includegraphics[width=0.5\textwidth]{figures/figure_lf-light.pdf}}
\caption{Lifeforms have different tolerances for sun and shade among British angiosperms. Each panel is the distribution of Ellenberg light indicator values on an integer scale of 1-9 for different Raunki\ae r life forms. Height of the bars indicate the raw proportion of species in each bin for that lifeform. The sample size for each lifeform is listed next in parentheses. The mean (open circle) and 95\% confidence intervals (black line) around the mean Ellenberg light indicator value for each lifeform based on phylogenetic regression are above the histogram.} 
\label{fig:lf-light}
\end{figure}

\begin{figure}[ht]
\centerline{\includegraphics[width=0.5\textwidth]{figures/figure_SR-lf.pdf}}
%\centerline{\includegraphics[angle=270]{figures/figure_SRmultReg.eps}} % Needed this to upload file to journal website
\caption{The effect of light on stomatal ratio depends on Raunki\ae r life form. Greater Ellenberg light indicator values (\el) are associated with greater stomatal ratio ($\mathrm{SR_{even}}$) in therophytes, hemicryptophytes, and chamaephytes but not geophytes and phanerophytes. The maximum likelihood slope from phylogenetic regression is given with statistical significance based on 10$^4$ parametric bootstrap samples. Numbers in parentheses next to Raunki\ae r life form are the sample sizes in the final dataset. Estimated slopes (solid line) and 95\% bootstrapped confidence intervals (gray polygon between dashed lines) are plotted against raw data. Points have been jittered for visual clarity.} 
\label{fig:SR-lf}
\end{figure}

@@

\begin{figure}[ht]
\centerline{\includegraphics{figures/figureS_violin.pdf}}
\caption{Most hydrophytes are hyperstomatous, having most stomata on the adaxial (`upper') surface (high $\mathrm{SD_{propAd}}$). The violin plot shows stomatal ratio as a function of Raunki\ae r lifeform. The width of the grey polygons indicates the density of data. Length of grey polygon indicate the range of the data; the point indicates the median; the thick lines indicate the 0.25 and 0.75 quantiles. Sample sizes per lifeform in the dataset are given above the upper plot margin. $\mathrm{SD_{ad}}$ and $\mathrm{SD_{total}}$ stand for the stomatal density on the adaxial surface and the total leaf surface (adaxial plus abaxial), respectively.}
\label{fig:violin}
\end{figure}

\begin{figure}[ht]
\centerline{\includegraphics[width=\textwidth]{figures/figure_phylo-habit.pdf}}
\caption{Phylogenetic diversification of stomatal ratio follows growth form and light tolerance. At the center is the phylogenetic tree for 372 species of British angiosperms. For each species, the green wedges indicate plant habit and the orange wedges indicate \el. The outer circle indicates the stomatal ratio for each species. As shown in the legend above, greater stomatal ratio means stomata are more evenly distributed across both leaf surfaces; lower stomatal ration means that most stomata are on the lower surface.} 
\label{fig:phylo-habit}
\end{figure}

\begin{figure}[ht]
\centerline{\includegraphics{figures/figureS_compare-growth-forms.pdf}}
\caption{Raunki\ae r lifeform and plant habit broadly overlap. The dot chart shows for each Raunki\ae r lifeform, the proportion that overlap with a given plant habit. For example, phanerophytes are mostly trees and shrubs, geophytes are all perennial, therophytes are mostly annuals, and so forth.}
\label{fig:compare-growth-forms}
\end{figure}

\begin{figure}[ht]
\centerline{\includegraphics[width=0.5\textwidth]{figures/figure_hf-light.pdf}}
\caption{Growth forms have different tolerances for sun and shade among British angiosperms. Each panel is the distribution of Ellenberg light indicator values on an integer scale of 1-9 for different growth forms. Height of the bars indicate the raw proportion of species in each bin for that lifeform. The sample size for each lifeform is listed next in parentheses. The mean (open circle) and 95\% confidence intervals (black line) around the mean Ellenberg light indicator value for each lifeform based on phylogenetic regression are above the histogram.} 
\label{fig:hf-light}
\end{figure}

\begin{figure}[ht]
\centerline{\includegraphics[width=0.5\textwidth]{figures/figure_SR-hf.pdf}}
%\centerline{\includegraphics[angle=270]{figures/figure_SRmultReg.eps}} % Needed this to upload file to journal website
\caption{The effect of light on stomatal ratio depends on growth form. Greater Ellenberg light indicator values (\el) are associated with greater stomatal ratio ($\mathrm{SR_{even}}$) in annual, biennual, and perennial herbs, but not shrubs or trees. The maximum likelihood slope from phylogenetic regression is given with statistical significance based on 10$^4$ parametric bootstrap samples. Numbers in parentheses next to growth form are the sample sizes in the final dataset. Estimated slopes (solid line) and 95\% bootstrapped confidence intervals (gray polygon between dashed lines) are plotted against raw data. Points have been jittered for visual clarity.} 
\label{fig:SR-hf}
\end{figure}
\end{document}

>>>>

$^1$ Biodiversity Research Centre and Botany Department, University of British Columbia, Vancouver, British Columbia V6T 1Z4, Canada \\
\\
\textit{Author for correspondence:} \\
\textit{Christopher D. Muir} \\
\textit{Tel:} +17782284851 \\
\textit{Email:} chrisdmuir@gmail.com \\
University of British Columbia \\
6270 University Blvd. \\
Vancouver, BC, Canada \\
V6T 1Z4 \\
\\
Short title: Shedding light on stomatal evolution\\
\\
Word count: \\
Summary: 199\\
Introduction: 1355\\
Materials and Methods: 1698\\
Results: 582\\
Discussion: 1935\\
Acknowledgement: 26\\
4 Figures and 1 Table, 5 Supplemental Figures

\linenumbers
\setstretch{\stretchy}
%\pagenumbering{gobble}
%\raggedright

%\clearpage

@@

\begin{itemize}

	\item In most plants, stomata are located only on the abaxial leaf surface (hypostomy), but many plants have stomata on both surfaces (amphistomy). High light and herbaceous growth form have been hypothesized to favor amphistomy, but these hypotheses have not been rigourously tested together using phylogenetic comparative methods.

	\item I leveraged a large dataset including stomatal ratio, Ellenberg light indicator value, growth form, and phylogenetic relationships for 372 species of British angiosperms. I used phylogenetic comparative methods to test how light and/or growth form influence stomatal ratio and density.

	\item High light and herbaceous growth form are correlated with amphistomy, as predicted, but they also interact; the effect of light is pronounced in therophytes (annuals) and perennial herbs, but muted in phanerophytes (shrubs and trees). Furthermore, amphistomy and stomatal density evolve together in response to light.

	\item Comparative analyses of British angiosperms reveal two major insights. First, light and growth form interact to shape stomatal ratio; amphistomy is common under high light, but mostly for herbs. Second, coordinated evolution of adaxial stomatal density and light tolerance indicates that amphistomy helps to optimally balance light acquisition with gas exchange. Stomatal ratio may have potential as a functional trait for paleoecology and crop improvement.

\end{itemize}

@@

In answering these questions, I both reassessed previous hypotheses using newer phylogenetic comparative methods and evaluated previously untested hypotheses. I predicted \textit{a priori} that light habitat and growth would be correlated. Species with faster life histories, especially therophytes (annuals), would on average inhabit sunnier environments than species with slower life histories, especially phanerophytes (shrubs and trees). Based on hypotheses from previous studies, I also predicted that herbaceous growth form and high light would be associated with amphistomy, even after controlling for phylogenetic nonindependence. Although these predictions have been tested previously, it is critical to reevaluate them here with updated methods because the subsequent untested hypotheses build on these results. The first novel hypothesis I tested predicts that light and growth form interact. Specifically, I hypothesized that both high light and herbaceous growth would be required to favor a more even stomatal ratio (i.e. amphistomy). Finally, I tested whether amphistomy is part of a coordinated syndrome of traits that promote higher photosynthetic rate. If high light and growth form favor amphistomy because it increases photosynthesis, then it follows that they should also favor other stomatal traits that reinforce this advantage. If evolved increases in stomatal ratio are mediated by shifting abaxial stomata to the adaxial surface, holding total stomatal density constant, then the overall increase in CO$_2$ diffusion would be small. In contrast, if amphistomy evolves by increasing adaxial stomatal density while holding abaxial density constant, then \textit{total} stomatal density must increase as well. Evolutionary coordination of amphistomy and high stomatal density would thus reinforce one another, increasing CO$_2$ supply to chloroplasts more than changes in either trait would in isolation. Understanding selection on coordinated traits can explain the evolution of major functional trait axes and syndromes.

@@

Overall, $\mathrm{SR_{even}}$ increased with \el, but there was a significant interaction ($\Delta\mathrm{AIC} > 2$, Table~\ref{table:srAIC}) between Raunki\ae r life form and \el~(Fig.~\ref{fig:SR-lf}). When classified based on plant habit, growth form interacted with \el~less strongly ($\Delta \textrm{AIC} = 2.4$; Fig.~\ref{fig:SR-hf}). Raunki\ae r life form explained variation in stomatal ratio better than habit (lower AIC; Table~\ref{table:srAIC}), therefore I focus hereafter on those analyses. Both life form and \el~significantly increased model fit, though \el~had a markedly larger effect on model AIC (Table~\ref{table:srAIC}). The significant interaction is caused by different slopes between life forms. Among life forms with the overall greatest \el~(therophytes, hemicryptophytes, and chamaephytes, see Fig.~\ref{fig:lf-light}), there was a strong positive relationship between \el~and $\mathrm{SR_{even}}$. Parametrically bootstrapped 95\% confidence intervals for the slope did not overlap zero (Fig.~\ref{fig:SR-lf}). The slope was weakly positive or not significantly different from zero in the most shade-adapted life forms (geophytes and phanerophytes), albeit the patterns were distinct in these groups. There were both hypostomatous ($\mathrm{SR_{even}} \approx 0$) and amphistomatous ($\mathrm{SR_{even}} \approx 1$) geophytes, but these were distributed across \els. In contrast, phanerophytes were nearly always hypostomatous regardless of \el.

@@

The weak or nonsignificant relationship between \el~and stomatal ratio in geophytes and phanerophytes suggests that in some cases amphistomy may not reliably indicate open habitat without further information. For example, perhaps amphistomatous geophytes from partially shaded habitats are spring ephemerals, so they  experience high light during their growth phase, but this has not been tested. Likewise, phanerophytes are almost always hypostomatous \citep[see also][]{Muir_2015}. Most British phanerophytes are tall, hypostomatous trees, but the exceptions are telling. For example, the most amphistomatous phanerophyte in this dataset is \textit{Brassica oleracea}, a short-statured biennial that has more in common physiologically with hemicryptophytes than other phanerophytes. The other amphistomatous phanerophytes in this data set (\textit{Populus nigra} and \textit{Lavatera arborea}) are fast-growing pioneer species.

Finally, phylogenetic information should improve inferences about paleoclimates because there is appreciable phylogenetic signal in stomatal ratio. The phylogenetic half-life of stomatal ratio evolution, after accounting for \el~and Raunki\ae r life form, is $\mathrm{log}(2) / \alpha = 1.5$ my (see Table~\ref{table:srAIC} for maximum likelihood estimates of $\alpha$, the return rate in the Ornstein-Uhlenbeck model of trait evolution). This lag time may indicate that evolving to the `optimum' is constrained by the shape of the fitness landscape \citep{Muir_2015} or that other unmeasured factors which affect stomatal ratio have some phylogenetic signal. Regardless of the mechanism, this fact means that researchers may be able to use data from closely related species to improve paleoenvironment reconstruction. Despite there being phylogenetic signal, residual phylogenetic variation in stomatal ratio at the broad phylogenetic scale encompassed by British angiosperms should be at stationarity. The observed variance in stomatal ratio, after accounting for \el~and Raunki\ae r life form, was indistinguishable from that expected for a trait at stationarity under an Ornstein-Uhlenbeck process (see Results). This may not be the case for younger clades that have radiated in the past few million years.

\subsection*{Coordinated evolution of stomatal ratio and density in response to light}

Variation in stomatal ratio is determined primarily by evolution of adaxial stomatal density and is coordinated with increases in total leaf stomatal density summed across both surfaces. Note here that I am referring only to evolutionary variation in stomatal ratio among species; different processes may mediate within species variation or plastic responses. Phylogenetic analyses show that changes in stomatal ratio and total stomatal density, especially in response to \el, are predominantly mediated by changes in adaxial stomatal density. To my knowledge, this highly nonrandom pattern among British angiosperms has not been demonstrated before, but it parallels evolutionary changes wrought by domestication \citep{Milla_etal_2013}; crop species tend to have higher adaxial stomatal density than their wild relatives.

@@

I refer to the second hypothesis as the `coordination' hypothesis. In this scenario, ecological conditions such as high light select for both increased total stomatal density and for amphistomy because these traits work well in coordination with one another. For example, if stomatal density were very high on a hypostomatous plant, then CO$_2$ would be more strongly limited by the mesophyll. Adding a second parallel pathway for diffusion by developing stomata on both surfaces would restore a more optimal balance between stomatal and mesophyll limitations. Conversely, there would be little benefit to amphistomy when total stomatal density is low because CO$_2$ diffusion is strongly limited by stomatal resistance, and therefore photosynthetic rate is not sensitive to changes in mesophyll diffusion mediated by stomatal ratio. A related prediction is that increased atmospheric CO$_2$ may select for reduced stomatal ratio and density primarily by decreasing adaxial stomatal density, but this has not been well tested \citep[but see][]{Woodward_Bazzaz_1988}. These results suggest that coordination between stomatal ratio and density might play a greater role than previously appreciated in optimizing CO$_2$ supply and demand under different light regimes \citep[see also][]{Beerling_Kelly_1996}.

@@

\begin{table}[ht]
 \setstretch{\stretchy}
  \topcaption{Interaction beween species' Ellenberg light indicator value (\el) and Raunki\ae r life form predict stomatal ratio ($\mathrm{SR_{even}}$). I compared phylogenetic linear models using the Akaike Information Criterion (AIC), where AIC = $2k - 2 \mathrm{log}(\mathcal{L})$. $k$ is the number of model parameters and $\mathcal{L}$ is the model likelihood. Given a set of candidate models, the difference in AIC between a model and the lowest AIC ($\Delta$AIC) indicates the relative fit of competing models. The correlation coefficient $r^2$ is another indicator of model fit. $\alpha$ and $\sigma^2$ are the return rate and diffusion parameters of the Ornstein-Uhlenbeck model of trait evolution. }
  \begin{center}
  \begin{tabular}{@{} l c c c c c c c @{}}
  \toprule
  Model: $\mathrm{SR_{even}} \sim$ &  $\alpha$ & $\sigma^2$ &  $r^2$ & $k$ &  log($\mathcal{L}$) &  AIC &  $\Delta$AIC \\
  \midrule  
    \el~$\times$ Raunki\ae r life form
  & 0.46
  & 0.068
  & 0.34
  & 12
  & -33.3
  & 90.6
  & 0 \\
    \el~$\times$ growth form
  & 0.46
  & 0.07
  & 0.32
  & 12
  & -38.2
  & 100.3
  & 9.8 \\
    \el~+ Raunki\ae r life form
  & 0.47
  & 0.072
  & 0.32
  & 8
  & -40.3
  & 96.5
  & 6 \\
     \el~+ growth form
  & 0.51
  & 0.08
  & 0.31
  & 8
  & -43.3
  & 102.7
  & 12.1 \\
    Raunki\ae r life form
  & 0.34
  & 0.067
  & 0.15
  & 7
  & -79.2
  & 172.4
  & 81.8 \\
    growth form
  & 0.35
  & 0.069
  & 0.13
  & 7
  & -82.5
  & 178.9
  & 88.4 \\
    \el
  & 0.64
  & 0.108
  & 0.26
  & 4
  & -59.3
  & 126.6
  & 36.1 \\
    null
  & 0.29
  & 0.067
  & 0
  & 3
  & -107.5
  & 221
  & 130.5 \\
	\bottomrule
	\end{tabular}
	\label{table:srAIC}
\end{center}
\end{table}

@@

\begin{figure}[ht]
\centerline{\includegraphics[width=\textwidth]{figures/figure_phylo-lifeform.pdf}}
\caption{Phylogenetic diversification of stomatal ratio follows growth form and light tolerance. At the center is the phylogenetic tree for 372 species of British angiosperms. For each species, the green wedges indicate Raunki\ae r life form and the orange wedges indicate \el. The outer circle indicates the stomatal ratio ($\mathrm{SR_{even}}$) for each species. As shown in the legend above, greater stomatal ratio means stomata are more evenly distributed across both leaf surfaces; lower stomatal ratio means that most stomata are on the lower surface.} 
\label{fig:phylo-lifeform}
\end{figure}

\begin{figure}[ht]
\centerline{\includegraphics[width=0.5\textwidth]{figures/figure_lf-light.pdf}}
\caption{Life forms have different tolerances for sun and shade among British angiosperms. Each panel is the distribution of Ellenberg light indicator values on an integer scale of 1-9 for different Raunki\ae r life forms. Height of the bars indicate the raw proportion of species in each bin for that life form. The sample size for each life form is listed next in parentheses. The mean (open circle) and 95\% confidence intervals (black line) around the mean Ellenberg light indicator value for each life form based on phylogenetic regression are above the histogram.} 
\label{fig:lf-light}
\end{figure}

\begin{figure}[ht]
\centerline{\includegraphics[width=0.5\textwidth]{figures/figure_SR-lf.pdf}}
%\centerline{\includegraphics[angle=270]{figures/figure_SR-lf.eps}} % Needed this to upload file to journal website
\caption{The effect of light on stomatal ratio depends on Raunki\ae r life form. Greater Ellenberg light indicator values (\el) are associated with greater stomatal ratio ($\mathrm{SR_{even}}$) in therophytes, hemicryptophytes, and chamaephytes but not geophytes and phanerophytes. The maximum likelihood slope from phylogenetic regression is given with statistical significance based on 10$^4$ parametric bootstrap samples. Numbers in parentheses next to Raunki\ae r life form are the sample sizes in the final dataset. Estimated slopes (solid line) and 95\% bootstrapped confidence intervals (gray polygon between dashed lines) are plotted against raw data. Points have been jittered for visual clarity.} 
\label{fig:SR-lf}
\end{figure}

@@

\begin{figure}[ht]
\centerline{\includegraphics{figures/figureS_violin.pdf}}
\caption{Most hydrophytes are hyperstomatous, having most stomata on the adaxial (`upper') surface (high $\mathrm{SD_{propAd}}$). The violin plot shows stomatal ratio as a function of Raunki\ae r life form. The width of the grey polygons indicates the density of data. Length of grey polygon indicate the range of the data; the point indicates the median; the thick lines indicate the 0.25 and 0.75 quantiles. Sample sizes per life form in the dataset are given above the upper plot margin. $\mathrm{SD_{ad}}$ and $\mathrm{SD_{total}}$ stand for the stomatal density on the adaxial surface and the total leaf surface (adaxial plus abaxial), respectively.}
\label{fig:violin}
\end{figure}

\begin{figure}[ht]
\centerline{\includegraphics{figures/figureS_compare-growth-forms.pdf}}
\caption{Raunki\ae r life form and plant habit broadly overlap. The dot chart shows for each Raunki\ae r life form, the proportion that overlap with a given plant habit. For example, phanerophytes are mostly trees and shrubs, geophytes are all perennial, therophytes are mostly annuals, and so forth.}
\label{fig:compare-growth-forms}
\end{figure}

\begin{figure}[ht]
\centerline{\includegraphics[width=\textwidth]{figures/figure_phylo-habit.pdf}}
\caption{Phylogenetic diversification of stomatal ratio follows growth form and light tolerance. At the center is the phylogenetic tree for 372 species of British angiosperms. For each species, the green wedges indicate plant habit and the orange wedges indicate \el. The outer circle indicates the stomatal ratio ($\mathrm{SR_{even}}$) for each species. As shown in the legend above, greater stomatal ratio means stomata are more evenly distributed across both leaf surfaces; lower stomatal ratio means that most stomata are on the lower surface.} 
\label{fig:phylo-habit}
\end{figure}

\begin{figure}[ht]
\centerline{\includegraphics[width=0.5\textwidth]{figures/figure_hf-light.pdf}}
\caption{Growth forms have different tolerances for sun and shade among British angiosperms. Each panel is the distribution of Ellenberg light indicator values on an integer scale of 1-9 for different plant habits. Height of the bars indicate the raw proportion of species in each bin for that habit. The sample size for each habit is listed next in parentheses. The mean (open circle) and 95\% confidence intervals (black line) around the mean Ellenberg light indicator value for each habit based on phylogenetic regression are above the histogram.} 
\label{fig:hf-light}
\end{figure}

\begin{figure}[ht]
\centerline{\includegraphics[width=0.5\textwidth]{figures/figure_SR-hf.pdf}}
%\centerline{\includegraphics[angle=270]{figures/figure_SR-hf.eps}} % Needed this to upload file to journal website
\caption{The effect of light on stomatal ratio depends on growth form. Greater Ellenberg light indicator values (\el) are associated with greater stomatal ratio ($\mathrm{SR_{even}}$) in annual, biennual, and perennial herbs, but not shrubs or trees. The maximum likelihood slope from phylogenetic regression is given with statistical significance based on 10$^4$ parametric bootstrap samples. Numbers in parentheses next to growth form are the sample sizes in the final dataset. Estimated slopes (solid line) and 95\% bootstrapped confidence intervals (gray polygon between dashed lines) are plotted against raw data. Points have been jittered for visual clarity.} 
\label{fig:SR-hf}
\end{figure}
\end{document}
